\vspace{-3em} \textbf{ Occultation mapping of Io's surface in the near-infrared \\ I: Inferring static maps}

### 2021-03-02 paper/paper.tex
Rodrigo's comments, a bunch of other changes

% Bibliography stuff
\bibliographystyle{aasjournal}

% Begin!
\begin{document}

% Title
\title{Occultation mapping of Io's surface in the near-infrared I: Inferring static maps}

% Author list
\author{Fran Bartoli\'c}
\email{fbartolic@flatironinstitute.org}
\affil{Center~for~Computational~Astrophysics, Flatiron~Institute, New~York, NY}
\affil{Centre for Exoplanet Science, SUPA, School of Physics and Astronomy, University of St. Andrews, St. Andrews, UK}
\author{Rodrigo Luger}
\author{Daniel Foreman-Mackey}
\affil{Center~for~Computational~Astrophysics, Flatiron~Institute, New~York, NY}
%

\begin{abstract}
Jupiter's moon Io is the most volcanically active body in the Solar System with hundreds of active volcanoes varying in intensity on different timescales.
Existing photometric observations in the near infrared taken during occultations by Jupiter and other Galilean moons encode a wealth of information about its surface.
These data are ideal for testing methods for inferring surface features of occulted bodies from time series observations.
We built a generative model of Io's surface using the code starry \href{https://rodluger.github.io/starry/}{\color{linkcolor}\faGithub} which enables fast analytic and differentiable computation of occultation light curves in emitted and reflected light.
    Our (fully Bayesian) model is able to recover known bright features on the surface of Io with only two light curves and without any assumptions on where the features are located or how many there are.
    This work is a first step towards constructing a model for time-variable surface features. 
    Although we do not present new findings about Io in this work, the methods we have developed are directly applicable to the problem of mapping stars and exoplanets.\href{https://github.com/fbartolic/volcano}{\color{linkcolor}\faGithub}
\end{abstract}

%
\section{Introduction}
The surface of Jupiter's moon Io is littered with hundreds of volcanoes which appear as bright spots in the near infrared and in intensity on timescales ranging from days to decades.
The volcanic activity is driven by tidal interactions with Jupiter and sustained by the Laplace resonance with Europa and Ganymede \citep{peale1979}.
Because of its extreme volcanic activity, Io is in many ways an analogue of a volcanically active exoplanet.
Volcanic exoplanets, often dubbed \emph{super-Ios} or \emph{lava worlds}, have gathered a lot of interest in recent years because both photometric and spectral signatures of volcanic activity on such worlds will likely be detectable in the near future with telescopes such as JWST and LUVOIR \citep{kaltenegger2010,henning2018,oza2019}.
Studying extreme volcanism on both Io and exoplanets is scientifically valuable because it provides a window into the properties of the early Earth and it informs theories of planet formation.

Existing exoplanet detections with potential volcanism include CoRoT-7b \citep{barnes2010}, the first rock exoplanet to have been discovered and one which likely heated by strong tidal forces; 55 Cancri e, whose inferred longitudinal offset in peak surface emission has been attributed to, among other things, lava flows on the surface \citep{demory2016,demory2016a,hammond2017}; several planets in the TRAPPIST-1 system in a Laplace-like resonance with likely volcanic activity \citep{kislyakova2017,dobos2019} and many others.
Not all exoplanets will have Io-like volcanism, some will have magma oceans due to their proximity to the star or volcanism caused by nuclear decay similar to early Earth volcanism.

Io has been observed extensively using both space and ground observatories.
High resolution images of Io's surface were taken by space missions such as Voyager \citep{smith1979}, Galileo \citep{belton1996} and Juno \citep{mura2020}.
It has also been resolved from the ground in the near infrared using disk resolved imaging
\citep{howell1985,simonelli1986,spencer1990} and adaptive optics observations \citep{marchis2000,marchis2005,dekleer2016}.
Most importantly for this work, Io has been sporadically observed for decades using high cadence near infrared photometry taken during occultations by Jupiter starting with \cite{spencer1990}.
Occultations by Jupiter occur twice every orbit lasting ~1.7 days and mutual occultations with Europa, Ganymede or Callisto happen every ~6 years when Earth pasess through the orbital plane of Galilean satellites.
The majority of occultations by Jupiter are observed when Io is in Jupiter's shadow ("in eclipse") while mutual occultations are almost always observed in sunlight because mutual occultations while Io is in eclipse are extraordinarily rare.
Only the brightest volcanoes are visible over the reflected light background when Io is illuminated by the Sun\citep{veeder1994,dekleer2016a}.

Both kinds of occultations have been used to study volcanic activity on the surface.
\cite{spencer1994} observed several occultations of Io by Europa and detected a major brightening of the most powerful of Io's volcanoes, Loki, relative to previous Voyager observations.
\cite{rathbun2002, rathbun2006,rathbun2010} used observations from NASA's IRTF telescope to study the long term variability of different volcanic spots finding evidence of periodicity in Loki's eruptions and established the transient nature of the volcanoes (only Loki is persistently active).
By studying an occultation of Io by Europa, \cite{dekleer2017} mapped the Loki region to a precision of about 2 kilometers.
In addition to observations of occultations in near infrared, several groups have been observing mutual occultations in the optical \citep[][and references therein]{arlot1974,saquet2018,morgado2016a} for decades with the purpose of inferring the optical albedo of Io and improving ephemeris precision for Galilean satellites.

None of the studies of occultations in the infrared attempted to infer a two-dimensional map of Io's surface.
The light curves are usually fit independently with the goal of inferring one-dimensional longitudinal variations in brightness on its surface.
Strong priors on the locations of individual volcanoes are assumed based on maps from space mission observations.
In this work, we build a fully probabilistic model in order to infer a two-dimensional \emph{map} of Io's volcanic surface from archival IRTF observations in the near infra red.
We assume that the map generating a given set of light curves is static (up to an overall amplitude) so we are restricted to modeling the surface during a timespan when the surface emission hasn't changed drastically.
In a subsequent paper we build on the model we have developed here and generalize to a case where the surface is time-variable.

The model is built on top of the recently developed code starry \citep[][Luger et al. 2020 in prep]{luger2019a} which enables fast analytic computation of occultation light curves and phase curves for objects whose surface features are represented in terms of spherical harmonics.
Starry can compute phase curves and occultations in both emitted light (for modeling the isotropic thermal emission from Io's surface) and reflected light (for mapping albedo variations).
It is many orders of magnitude faster and more accurate than pixel based codes and it computes exact gradients with respect to all parameters through automatic differentiation.
Starry also comes with extensive tools for visualizing spherical harmonic maps and simulating data.

Our goal for this paper is to develop a general model for mapping surfaces of occulted bodies with sparse high contrast features and to test it on a real world dataset. 
Because we have lots of prior information on what the surface of Io should look like, we can compare our inferences to known features on its surface and we can test the sensitivity of the inferred maps to different choices of priors and noise models.
The work should also be of interest to the Io research community as it presents a very different approach to modeling Io occultations without having to make strong assumptions on the locations of the hot spots.

The paper is organized as follows.
In \S\ref{sec:data} we describe the IRTF light curves of occultations by Jupiter which we use to infer maps.
In \S\ref{sec:model} we discuss the generative model for the data and we write down the relevant likelihood functions.
In \S\ref{sec:inverse_problem} we focus on the question of how to do Bayesian inference given the forward model specifications defined in \S\ref{sec:model}.
We discuss and test different priors on map features, the information content of light curves and fitting the modelsusing Hamiltonian Monte Carlo (HMC). 
We show results on simulated data.
\S\ref{sec:results} shows the results on real IRTF data for two pairs of light curves (closely spaced in time)
In particular, the inferred dynamic map, the  different component maps and accompanying coefficients, the uncertainties of the maps, sensitivity to different choices of priors, comparison of inferred locations and intensities of spots to known values from the literature.
Finally, in \S\ref{sec:exoplanets} we discuss applications of the model to future exoplanet obervations and test the model predictions on simulated JWST observations of a volcanic Super-Earth.

@@

\begin{figure}[h!]
    \begin{centering}
    \includegraphics[width=0.5\linewidth]{figures/irtf_max_flux.pdf}
    \oscaption{irtf_dataset}{%
        Maximum flux for each IRTF light curve of Io observed during an occultation of Io by Jupiter.
       \label{fig:irtf_max_flux}
    }
    \end{centering}
\end{figure}

The dataset we use to infer maps of Io's surface consists of 112 observations of Io in the near infrared taken from 1996 until 2018 using different instruments at NASA's IRTF observatory, the NSFCam  1-5 $\mu m$ camera \citep{shure1994} (TODO: WHICH FILTERS), the SpeX 0.8-5.5 $\mu m$ spectrographer/imager \citep{rayner2003} (TODO: DETAILS) and the iShell 1.1-5.3 $\mu m$  spectrographer/imager \citep{johnrayner2016}.
Although the observations span decades, the cadence is non-uniform, there are only a few observations taken between 2008 and 2016.
All of the observations were taken while Io was \emph{in eclipse}, meaning that it was in Jupiter's shadow and all of the observed emission from the surface is due to thermal radiation.
Observations of Io \emph{in sunlight} on the other hand probe both the thermal (volcanic) emission and the surface albedo variations in the near infrared.
Figure~\ref{fig:irtf_max_flux} shows the maximum flux in units of $\mathrm{GW}/\mu \mathrm{m}/\mathrm{sr}$ (TODO: UNDERSTAND THESE UNITS) for each light curve plotted as a function of time.
This value is approximately equal to the total flux emitted from the Jupiter facing side of Io.
The baseline brightness is varying stochastically by a large amount and there are several notable events of intense volcanic activity.

\begin{figure}[h!]
    \begin{centering}
    \includegraphics[width=0.5\linewidth]{figures/irtf_sample_lightcurves.pdf}
    \oscaption{irtf_dataset}{%
        A selection of sample light curves taken during occultations of Io by Jupiter in our dataset.
        The step-like morphology of the light curves is due to bright volcanoes on Io's surface coming in or out of view during an occultation.
        These light curves thus visibly encode information about the features on the surface.
        \label{fig:irtf_sample_lightcurves}
    }
    \end{centering}
\end{figure}

In Figure~\ref{fig:irtf_sample_lightcurves} we display a random subset of all the light curves
in the dataset.
All occultations last for $\sim4$ min.
The exposure cadence for each light curve is around a second with some variation between light curves.
It follows that over the course of a single exposure Jupiter's limb passes over about 15km on Io's surface which provides a lower bound for the size of the features that we can reliably estimate.
The shapes of light curves strongly deviate from the smooth variability one would expect assuming uniform thermal surface emission.
Especially prominent are light curves with clear step-like features which are present when bright localized spots come in or out of view during the course of an occultation.
The fact that these features are so clearly visible means that even individual light curves encode precise information on the longitudinal position of the spots.

The photometric quality of the light curves varies from year to year because multiple instruments were used over the years. 
As with all ground based photometry, the observations are influenced by atmospheric variability which results in some correlated noise in the light curves and because of this the flux is not always monotonically increasing or decreasing as one would expect.
The light curves lack errorbars. 
To estimate the errorbars we do some basic pre-processing on the light curves. 
We filter each light curve with the Savitzky-Golay filter as implemented in \textsf{SciPy} \citep{virtanen2020} with
a third order polynomial and a window length equal to 5\% of the number of data points and rounded up to the nearest odd integer. 
After filtering, we compute the residuals with respect to the filtered model and run iterative sigma-clipping, removing all data points 5 median absolute deviations (MAD) from the median value. 
Finally, we estimate the errorbars by computing the standard deviation of the sigma clipped residuals.
This estimate is far from perfect which is why we introducing a rescaling parameter when fitting the model.

\section{Model}
\label{sec:model}
\subsection{Orbital parameters}
\label{ssec:orbital_parameters}
To model the occultations we need to know the projected relative position of Jupiter with respect to Io and the orientation of Io at any given time.
To compute the geometry, we use the \href{https://ssd.jpl.nasa.gov/horizons.cgi}{JPL Horizons database} and the latest planetary ephemeris DE430 \citep{folkner2014} which is accurate to a few kilometers on Io's surface.
For both Jupiter and Io we need the Right Ascension and Declination and the angular (equatorial) diameter (\textsf{Ang-diam}).
In addition, we need to fix the orientation of Io we need the longitude and the latitude at the center of Io's disc as seen from Earth (\textsf{Ob-lon}, \textsf{Ob-lat}) and the counterclockwise angle between the Celestial North Pole unit vector projected onto the plane of the sky and the Io's north pole (\textsf{NP.ang}).
All longitudes are positive  in the direction of West.
\textsf{HORIZONS} provides all ephemeris with a minimum cadence of 1 second so we interpolate all values so that we can evaluate them for arbitrary times.

The coordinate system in \textsf{starry} is defined to be right handed such that the $\hat{z}$
axis points towards the observer and the $\hat{x}$ axis points to the right on the plane of the sky.
The radius of the occulted sphere is fixed to 1 and the orientation is specified by three angles,
the counterclockwise obliquity angle \textsf{obl} between the $\hat{y}$ axis and the North Pole of the sphere, the inclination angle \textsf{inc} which is set to $90^\circ$ if the North Pole is aligned with the $\hat{y}$ axis and the phase angle \textsf{theta} which rotates the sphere around the $\hat{y}$ axis in the eastward direction.
We have $\mathrm{obl}=\textsf{NP.ang}$, $\mathrm{inc}=90^\circ-\textsf{Ob-lat}$ and
$\mathrm{theta}=\textsf{Ob-lon}$.
The occultor position relative to the occulted object is given by
\begin{align}
    \mathrm{xo}/\gamma&=-\Delta\alpha\,\cos\delta\\
    \mathrm{yo}/\gamma&=\Delta\delta\\
    \mathrm{zo}/\gamma&=1
\end{align}
where $\Delta\alpha$ and $\Delta\delta$ are differences in Right Ascension and Declination respectively relative to the occulted object and $\gamma$ is the angular radius of the occulted sphere.

\subsection{Jupiter's effective radius}
Although the sky position of Jupiter relative to Io is known to a precision of a few kilometers, several complications arise when attempting to compute Jupiter's radius.
First, Jupiter is not spherical and its equatorial radius is greater than the polar radius by thousands of kilometers. 
Because \textsf{starry} does not support occultations by non-spherical occultors and because the effect is negligible at the resolution of our maps, we assume that Jupiter is locally spherical at the point of an occultation. 
We estimate an effective radius from the measurements of Jupiter's shape from Voyager radio occultation data \citep{lindal1981}.
These data come in the form of a plot of effective radius  as a function of Jupiter's planetocentric latitude at a fixed pressure of 100 mbar \citep[Fig.~7 in ][]{lindal1981}.
When computing the occultation latitude we fix Jupiter's latitude to the value at the center of Io's disc because 
the variation of Jupiter's effective radius along Io's disc is negligible.

Second, Jupiter is gaseous so it does not have a well defined boundary.
In principle, we should compute an effective radius of Jupiter at different locations (pressure) in the atmosphere and model an occultation of Io by a fuzzy occultor.
Although this is possible with Starry, it is unnecessary for our models because the characteristic scale height of Jupiter is around 27 km (CITE) which is below the uncertainty of our inferred maps.
Instead, we follow the approach in \cite{spencer1990} and compute the effective radius of Jupiter at about 2.2 mbar, the pressure (and the associated effective radius) at which a  bright source on the surface of Io fades by 50\% due to differential refraction during the course of an occultation.

Third, the information on the effective radius in \citep{lindal1981} is provided only at a fixed pressure of 100 mbar.
To adjust the values for a lower pressure of 2.2 mbar we assume an exponential pressure profile $P=P_0\, e^{-\Delta r/H}$ where $H$ is the scale height and $\Delta r$ is the height difference between the two pressure levels.
It follows that to convert the shape profile at 100 mbar to 2.2 mbar we need to add the factor
$-H\ln(2.2/100)$ which is assumed to be constant in the $\pm 21$ deg latitude range in which the occultations occur.
In addition to refractive absorption in Jupiter's atmosphere there is a also slight additional molecular absorption due to methane which \cite{spencer1990} estimate to be equal to around 12\% in their filter, we choose to ignore this because it is far below the resolution of our maps.

Finally, we have to account for the fact that the light from Io is getting significantly bent at the point of half reflective intensity in Jupiter's atmosphere which results in a smaller shadow of Jupiter on the surface of Io then would be the case for a straight propagation.
It is zero at the beginning of the disappearance of a hot spot when there is no refraction, increasing to one scale height $H$ at the half intensity point and then increasing further to a large value when Io disappears behind Jupiter's limb.
We ignore the variation in the bending and adopt a fixed value of one scale height for this effect which we subtract from the value of the effective radius.

In summary, to compute an effective radius of Jupiter for a given occultation of Io first we have to compute Jupiter's planetocentric latitude at which Io's disk disappears or reappears behind the limb, then use a modified shape profile from \cite{lindal1981} to get an effective radius and finally we have to subtract one scale height due to light bending.
There are substantial uncertainties in each of these steps.
The data about on the shape profile are quite old and it is not known if the structure of Jupiter's atmosphere stayed the same since 1980s. 
The shape profile also depends on the wind velocity structure and temperature which we do not take into account.
In addition to uncertainties about the atmospheric structure, there are uncertainties associated with digitizing the data shown in Fig.~7 in \cite{lindal1981} because it is not available in table form.
We digitized the figure by reading off the data points "by eye" and then interpolating them using cubic spline interpolation.

>>>>

% Bibliography stuff
\bibliographystyle{aasjournal}
\setlength\LTcapwidth{\textwidth}
\begin{document}

% Title
\title{
    \vspace{-3em}
\textbf{    Occultation mapping of Io's surface in the near-infrared \\ I: Inferring static maps} 
}
% Author list
\author{Fran Bartoli\'c}
\email{fb90@st-andrews.ac.uk}
\affil{Centre for Exoplanet Science, SUPA, School of Physics and Astronomy, University of St. Andrews, St. Andrews, UK}
\affil{Center~for~Computational~Astrophysics, Flatiron~Institute, New~York, NY}
\author{Rodrigo Luger}
\author{Daniel Foreman-Mackey}
\affil{Center~for~Computational~Astrophysics, Flatiron~Institute, New~York, NY}
%

\begin{abstract}
Jupiter's moon Io is the most volcanically active body in the Solar System with hundreds of active volcanoes varying in intensity on different timescales.
Existing photometric observations in the near infrared taken during occultations by Jupiter and other Galilean moons encode a wealth of information about its surface.
These data are ideal for testing methods for inferring surface features of occulted bodies from time series observations.
We built a generative model of Io's surface using the code starry \href{https://rodluger.github.io/starry/}{\color{linkcolor}\faGithub} which enables fast analytic and differentiable computation of occultation light curves in emitted and reflected light.
    Our fully Bayesian model is able to recover known bright features on the surface of Io using only two light curves and without any assumptions on the locations, shapes or the number of features.
    The methods we have developed are also directly applicable to the problem of mapping stars and exoplanets.\href{https://github.com/fbartolic/volcano}{\color{linkcolor}\faGithub}
\end{abstract}

%
\section{Introduction}
The surface of Jupiter's moon Io is covered with hundreds of volcanoes which appear as bright spots in the near infrared, their intensities vary on timescales ranging from days to decades.
The global heat flow on its surface is about 40 times larger than Earth's \citep{breuer2007,davies2010}
and about 50\% of the heat flow emanates from only 1.2\% of Io's surface \citep{veeder2012}.
This intense volcanic activity cannot be explained by radioactive decay which is what drives Earth's volcanism, instead it is driven by tidal interactions with Jupiter and sustained by the Laplace resonance with Europa and Ganymede \citep{peale1979}.
Contrary to Earth, Io lacks plate tectonics.
The heat transport mechanism which likely on Io is volcanism through the litosphere (the outermost shell of a terrestrial body), a process in which magma is transported to the surface by localised vents \cite{oreilly1981}.
Heat-pipe volcanism might have occurred in the early history of terrestrial planets in the Solar System \cite{moore2013,moore2017}, most notably on early Earth prior to the onset of plate tectonics.

Besides providing a window into early volcanic activity in the Solar System, Io is also in many ways an analogue of a volcanically active exoplanet.
Volcanic exoplanets, often dubbed \emph{super-Ios} or \emph{lava worlds}, have gathered a lot of interest in recent years because both photometric and spectral signatures of volcanic activity on such worlds will likely be detectable in the near future with telescopes such as JWST and LUVOIR \citep{kaltenegger2010,henning2018,oza2019,chao2020}.
Existing exoplanet detections with potential volcanism include CoRoT-7b \citep{barnes2010}, the first rocky exoplanet to have been discovered and one which is likely heated by strong tidal forces; 55 Cancri e, whose inferred longitudinal offset in peak surface emission has been attributed to, among other things, lava flows on the surface \citep{demory2016,demory2016a,hammond2017}; several planets in the TRAPPIST-1 system in a Laplace-like resonance with likely volcanic activity \citep{kislyakova2017,dobos2019} and many others.
Not all such exoplanets will have Io-like volcanism; some may have magma oceans due to their proximity to the star or volcanism caused by nuclear decay, similar to volcanism on present day Earth.

Io has been observed extensively using both space and ground observatories.
High resolution images of Io's surface were taken by space missions such as Voyager \citep{smith1979}, Galileo \citep{belton1996} and Juno \citep{mura2020}.
It has also been resolved from the ground in the near infrared using disk resolved imaging
\citep{howell1985,simonelli1986,spencer1990} and adaptive optics observations \citep{marchis2000,marchis2005,dekleer2016}.
Most importantly for this work, Io has been sporadically observed for decades using high cadence near infrared photometry taken during occultations by Jupiter starting with \cite{spencer1990}.
Occultations by Jupiter occur twice every orbit of ~1.7 days.
Mutual occultations with Europa, Ganymede or Callisto happen at different times over the course of a single year but this only happens every ~6 years or so when Earth passes through the orbital plane of the Galilean satellites.
The majority of occultations by Jupiter are observed when Io is in Jupiter's shadow (``in eclipse") while mutual occultations are almost always observed in sunlight.
Only the brightest volcanoes are visible over the reflected light background when Io is illuminated by the Sun \citep{veeder1994,dekleer2016a}.

Both kinds of occultations have been used to study volcanic activity on the surface.
\cite{spencer1994} observed several occultations of Io by Europa and detected a major brightening of the most powerful of Io's volcanoes, Loki, relative to previous Voyager observations.
\cite{rathbun2002},\citet{rathbun2006} and \citet{rathbun2010} used observations from NASA's IRTF telescope to study the long term variability of different volcanic spots, finding evidence of periodicity in Loki's eruptions and establishing the transient nature of the volcanoes (only Loki is persistently active).
By studying a (spatially resolved) occultation of Io by Europa, \cite{dekleer2017} mapped the Loki region to a precision of about 2 kilometers.
In addition to observations of occultations in the near infrared, several groups have been observing mutual occultations in the optical \citep[][and references therein]{arlot1974,saquet2018,morgado2016a} for decades with the purpose of inferring the optical albedo of Io and improving ephemeris precision for Galilean satellites.
Understanding the detailed albedo distribution is crucial to constraining the ephemeris to a very high precision.

None of the studies of occultations in the infrared attempted to infer a two-dimensional map of Io's surface.
The light curves are usually fit independently with the goal of inferring one-dimensional longitudinal variations in brightness on its surface, with
strong priors on the locations of individual volcanoes based on maps from space mission observations.
In this work, we build a fully probabilistic model in order to infer a two-dimensional \emph{map} of Io's volcanic surface from archival IRTF observations in the near infrared.
We assume that the map corresponding to a given set of light curves is static (up to an overall amplitude) so we are restricted to modeling the surface during a timespan when the surface emission hasn't changed drastically.
In a subsequent paper we build on the model we have developed here and generalize to a case where the surface is time-variable.

The model is built on top of the recently developed code \textsf{starry}\citep[][Luger et al. 2021 in prep]{luger2019a} which enables fast analytic computation of occultation light curves and phase curves for objects whose surface features are represented in terms of spherical harmonics.
Starry can compute phase curves and occultations in both emitted light (for modeling the isotropic thermal emission from Io's surface) and reflected light (for mapping albedo variations).
It is many orders of magnitude faster and more accurate than pixel based codes and it computes exact gradients with respect to all parameters through automatic differentiation.
\textsf{starry} also comes with extensive tools for visualizing spherical harmonic maps and simulating data.

Our goal for this paper is to develop a general model for mapping surfaces of occulted bodies with sparse, high contrast features and to test it on a real world dataset. 
Because we have lots of prior information on what the surface of Io should look like, we can compare our inferences to known features on its surface and test the sensitivity of the inferred maps to different choices of priors and noise models.
The work should also be of interest to the Io research community as it presents a very different approach to modeling Io occultations without having to make strong assumptions on the locations of the hot spots.

The paper is organized as follows.
In \S\ref{sec:data} we describe the IRTF light curves of occultations by Jupiter that we use to infer maps.
In \S\ref{sec:model} we discuss the generative model for the data and we write down the relevant likelihood functions.
In \S\ref{sec:inverse_problem} we focus on the question of how to do Bayesian inference given the forward model specifications defined in \S\ref{sec:model}.
We discuss and test different priors on map features, the information content of light curves and fitting the models using Hamiltonian Monte Carlo (HMC). 
We then show results on simulated data.
In \S\ref{sec:results} we show the results on real IRTF data for two pairs of light curves, one pair observed in 1998 and another in 2017.
We show the inferred maps and the parameters quantifying the location and intensity of the hot spots, we discuss the sensitivity of the results to different choices of priors and compare the inferred values to previous results from the literature.
Finally, in \S\ref{sec:discussion} we summarize the main results of the paper and discuss potential applications of our model to observations of volcanic exoplanets.

TODO: say that this paper is first in series of awesome papers.

\vspace{1em}

All of the figures in this paper were auto-generated
using the Azure Pipelines continuous integration (CI) service.
Icons next to each of the figures \codeicon \,
link to the exact script used to generate them to ensure the reproducibility
of our results.

@@

\begin{figure}[h!]
    \begin{centering}
    \includegraphics[width=\linewidth]{figures/irtf_max_flux.pdf}
    \oscaption{irtf_dataset}{%
        Maximum flux for each IRTF light curve of Io observed during an occultation of Io by Jupiter.
       \label{fig:irtf_max_flux}
    }
    \end{centering}
\end{figure}

The dataset we use to infer maps of Io's surface consists of 112 sets of observations of Io in the near infrared taken from 1996 until 2018 using different instruments at NASA's IRTF observatory, the NSFCam  1-5 $\mu m$ camera \citep{shure1994} (TODO: WHICH FILTERS), the SpeX 0.8-5.5 $\mu m$ spectrographer/imager \citep{rayner2003} (TODO: DETAILS) and the iShell 1.1-5.3 $\mu m$  spectrographer/imager \citep{johnrayner2016} (TODO: fix citation).
Although the observations span decades, the cadence is non-uniform, there are only a few observations taken between 2008 and 2016.
All of the observations were taken while Io was \emph{in eclipse}, meaning that it was in Jupiter's shadow and all of the observed emission from the surface is due to thermal radiation.
Observations of Io \emph{in sunlight} on the other hand probe both the thermal (volcanic) emission and the surface albedo variations in the near infrared.
Figure~\ref{fig:irtf_max_flux} shows the maximum flux for each occultation light curve in units of $\mathrm{GW}/\mu \mathrm{m}/\mathrm{sr}$ as a function of time.
This value is approximately equal to the total flux emitted from the Jupiter-facing side of Io.
The baseline brightness is varying stochastically by a large amount and there are several notable events of intense volcanic activity.

\begin{figure}[h!]
    \begin{centering}
        \includegraphics[width=\linewidth]{figures/irtf_sample_lightcurves.pdf}
    \oscaption{irtf_dataset}{%
        A selection of sample light curves taken during occultations of Io by Jupiter in our dataset.
        The step-like morphology of the light curves is due to bright volcanoes on Io's surface coming in or out of view during an occultation.
        These light curves thus visibly encode information about the features on the surface.
        \label{fig:irtf_sample_lightcurves}
    }
    \end{centering}
\end{figure}

In Figure~\ref{fig:irtf_sample_lightcurves} we display a random subset of all the light curves
in the dataset. 
Each light curve covers a timespan of $\sim 4$ minutes which is the duration of the ingress and ingress parts of the entire occultation.
The exposure cadence for each light curve is about a second, with some variation between light curves.
It follows that over the course of a single exposure Jupiter's limb passes over about 15km on Io's surface which provides a lower bound for the size of the features that we can reliably estimate.
The shapes of light curves strongly deviate from the smooth variability one would expect assuming uniform thermal surface emission.
Especially prominent are light curves with clear step-like features which are present when bright localized spots come in or out of view during the course of an occultation.
The fact that these features are so clearly visible means that even individual light curves encode precise information on the longitudinal position of the spots.

The photometric quality of the light curves varies from year to year because multiple instruments were used over the years. 
As with all ground-based photometry, the observations are influenced by atmospheric variability which results in some correlated noise in the light curves, because of this the flux is not always monotonically increasing or decreasing as one would expect.
Complicating matters, the light curves lack errorbars. 
We deal with this by treating all errobars as free parameters with a common characteristic scale in the final model.

\section{Model}
\label{sec:model}
\subsection{Orbital parameters}
\label{ssec:orbital_parameters}
To model the occultations we need to know the projected relative position of Jupiter with respect to Io and the orientation of Io at any given time.
To compute the geometry, we use the \href{https://ssd.jpl.nasa.gov/horizons.cgi}{JPL Horizons database} and the latest planetary ephemeris DE430 \citep{folkner2014} which is accurate to a few kilometers on Io's surface.
For both Jupiter and Io we need the right ascension and declination and the angular (equatorial) diameter (\textsf{Ang-diam}).
In addition, we need to fix the orientation of Io we need the longitude and the latitude at the center of Io's disc as seen from Earth (\textsf{Ob-lon}, \textsf{Ob-lat}) and the counterclockwise angle between the celestial north pole unit vector projected onto the plane of the sky and the Io's north pole (\textsf{NP.ang}).
All longitudes are positive  in the direction of west.
\textsf{HORIZONS} provides all ephemeris with a minimum cadence of 1 second so we interpolate all values so that we can evaluate them for arbitrary times.

The coordinate system in \textsf{starry} is defined to be right handed such that the $\hat{z}$
axis points towards the observer and the $\hat{x}$ axis points to the right on the plane of the sky.
The radius of the occulted sphere is fixed to 1 and the orientation is specified by three angles,
the counterclockwise obliquity angle \textsf{obl} between the $\hat{y}$ axis and the north pole of the sphere, the inclination angle \textsf{inc} which is set to $90^\circ$ if the north pole is aligned with the $\hat{y}$ axis and the phase angle \textsf{theta} which rotates the sphere around the $\hat{y}$ axis in the eastward direction.
We have $\mathrm{obl}=\textsf{NP.ang}$, $\mathrm{inc}=90^\circ-\textsf{Ob-lat}$ and
$\mathrm{theta}=\textsf{Ob-lon}$.
The occultor position relative to the occulted object is given by
\begin{align}
    \mathrm{xo}/\gamma&=-\Delta\alpha\,\cos\delta\\
    \mathrm{yo}/\gamma&=\Delta\delta\\
    \mathrm{zo}/\gamma&=1
\end{align}
where $\Delta\alpha$ and $\Delta\delta$ are differences in right ascension and declination respectively relative to the occulted object and $\gamma$ is the angular radius of the occulted sphere.

\subsection{Jupiter's effective radius}
\label{ssec:effective_radius}
Although the sky position of Jupiter relative to Io is known to a precision of a few kilometers, several complications arise when attempting to compute Jupiter's radius.
First, Jupiter is not spherical and its equatorial radius is greater than the polar radius by thousands of kilometers. 
Because \textsf{starry} does not (yet) support occultations by non-spherical occultors and because the effect is negligible at the resolution of our maps, we assume that Jupiter is locally spherical at the point of an occultation. 
We estimate an effective radius from the measurements of Jupiter's shape from Voyager radio occultation data \citep{lindal1981}.
These data come in the form of a plot of effective radius  as a function of Jupiter's planetocentric latitude at a fixed pressure of 100 mbar \citep[Fig.~7 in ][]{lindal1981}.
When computing the occultation latitude we fix Jupiter's latitude to the value at the center of Io's disc because 
the variation of Jupiter's effective radius along Io's disc is negligible.

Second, Jupiter is gaseous so it does not have a well defined boundary.
In principle, we should compute an effective radius of Jupiter at different altitudes (pressure) in the atmosphere and model an occultation of Io by a fuzzy occultor.
Although this is possible with \textsf{starry}, it is unnecessary for our models because the characteristic scale height of Jupiter is around 27 km (CITE) which is below the uncertainty of our inferred maps.
Instead, we follow the approach in \cite{spencer1990} and compute the effective radius of Jupiter at about 2.2 mbar, the pressure (and the associated effective radius) at which a  bright source on the surface of Io fades by 50\% due to differential refraction during the course of an occultation.

Third, the information on the effective radius in \cite{lindal1981} is provided only at a fixed pressure of 100 mbar.
To adjust the values for a lower pressure of 2.2 mbar we assume an exponential pressure profile $P=P_0\, e^{-\Delta r/H}$ where $H$ is the scale height and $\Delta r$ is the height difference between the two pressure levels.
It follows that to convert the shape profile at 100 mbar to 2.2 mbar we need to add the factor
$-H\ln(2.2/100)$ which is assumed to be constant in the $\pm 21$ deg latitude range in which the occultations occur.
In addition to refractive absorption in Jupiter's atmosphere there is a also slight additional molecular absorption due to methane which \cite{spencer1990} estimate to be equal to around 12\% in their filter, we choose to ignore this because it is far below the resolution of our maps.

Finally, we have to account for the fact that the light from Io is getting significantly bent at the point of half reflective intensity in Jupiter's atmosphere which results in a smaller projected limb of Jupiter on the surface of Io then would be the case for straight propagation.
It is zero at the beginning of the disappearance of a hot spot when there is no refraction, increasing to one scale height $H$ at the half intensity point and then increasing further to a large value when Io disappears behind Jupiter's limb.
We ignore the variation in the bending and adopt a fixed value of one scale height for this effect which we subtract from the value of the effective radius.

In summary, to compute an effective radius of Jupiter for a given occultation of Io first we have to compute Jupiter's planetocentric latitude at which Io's disk disappears or reappears behind the limb, then use a modified shape profile from \cite{lindal1981} to get an effective radius and finally we have to subtract one scale height due to light bending.
There are substantial uncertainties in each of these steps.
The data on the shape profile are quite old and it is not known if the structure of Jupiter's has remained constant since the 1980s. 
The shape profile also depends on the wind velocity structure and temperature which we do not take into account.
In addition to uncertainties about the atmospheric structure, there are uncertainties associated with digitizing the data shown in Fig.~7 in \cite{lindal1981} because it is not available in table form.
We digitized the figure by reading off the data points ``by eye" and then interpolating them using cubic spline interpolation.
